Binarity's Biases on Transit Survey Occurrence Rates

### 2018-02-23 paper/main.tex
simple model jnw edits

Since the effects of binarity are most pronounced when the two stellar
components are similar, we begin by considering a universe in which
all stars are identical, with mass $M$, radius $R$, and luminosity
$L$.  Some fraction of them exist in twin binaries. All planets have
size $r\p$.  In this scenario, stellar radii are never misclassified
because the combined light of a binary has the same color and spectrum
as a single star.  We further assume that planets occur around single
stars and members of binaries at the same rate, $\Lambda(r\p)$.

Our naive observers detect transiting planets with two different
depths. Since 
\begin{equation}
    \delta = \left(\frac{r\p}{R}\right)^2 \frac{L}{L_{\rm tot}},
    \label{eq:delta_obs_general} 
\end{equation}
one class of transit signal, with depth $(r\p/R)^2$, occurs in single
star systems. The other class, with half the depth, occurs in twin
binary systems, leading to apparent planet radii $\sqrt{2}$ smaller
than the true radii.

To calculate the occurrence rate of each planet type, our observers
identify the ``stars'' from their survey that appear to have been
searched for signals of each depth.  These transits are detected
within a certain maximum
distance~\citep[see][]{pepper_using_2003,pepper_searching_2005}, which
from Equation~\ref{eq:S_N_thresh} scales as
\begin{equation}
  d_{\rm max} \propto \delta \, L_{\rm tot}^{1/2}.
  \label{eq:dmax}
\end{equation}
Assuming that stars are uniformly distributed in space, the number of
searchable stars $N_\star$ is proportional to
\begin{equation}
    N_\star \propto n \delta^3 L_{\rm tot}^{3/2},
\label{eq:N_searchable_prop}
\end{equation}
where $n$ is the number per unit volume of single or binary star
systems.

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{figures/visualize_volumes.pdf}
    \end{center}
    \caption{
        Searchable volumes for single stars ({\it left}) and twin
        binaries ({\it right}), assuming that all stars have the same
        mass, radius, and luminosity, and that all planets have the
        same radius $r\p$.
        {\it Top:} At an apparent radius $r\a=r\p$, the
        binarity-ignoring observers search twin binaries out to a
        distance $\sqrt{2}$ times greater than they search single
        stars.  Because the transit depths are diluted, there are no 
        planets in twin binaries with $r\a=r\p$.
        {\it Bottom:} At $r\a=r\p/\sqrt{2}$, the observed transit
        depths are smaller by a factor of two, so the maximum
        searchable distances are half those at $r\a=r\p$.  The only
        detected planets with $r\a=r\p/\sqrt{2}$ orbit twin binaries.
    }
    \label{fig:model_1_volumes}
\end{figure}

When selecting sources that appear to have been searched, the
observers admit some binaries.  Since the binaries are twice as
luminous as singles, at fixed $\delta$ they are included out to a
distance that is larger by a factor of $\sqrt{2}$.  To detect the
half-strength signal, the observers need the noise level to be lower
by a factor of two, which means the flux must be four times higher
than that of a single star at distance $d_0$.  This is true of single
stars within a distance $d_0/2$, and binaries within a distance
$d_0\sqrt{2}/2$.  The searchable volumes are illustrated in
Figure~\ref{fig:model_1_volumes}.

Putting it all together, the {\it apparent} occurrence rate
$\Lambda\a$ of planets of radius $r\p$ will be
\begin{equation}
    \Lambda\a(r\p) = 
        \frac{\Lambda(r\p) \, n\s d_0^3}
        {n\s d_0^3 + n\b (d_0\sqrt{2})^3}
    \label{eq:rate_at_rp}
\end{equation}
where $n\s$ and $n\b$ are the number densities of single and binary
star systems, respectively.  This apparent rate is larger than the
true rate by a factor
\begin{equation}
    \frac{\Lambda\a(r\p)}{\Lambda(r\p)} = 
        \frac{1}{1 + 2^{3/2}(n\b/n\s)}.
    \label{eq:correction_rp}
\end{equation}
The observers will also claim to have discovered a class of planet
with radius $r\p/\sqrt{2}$ and an occurrence rate given by
\begin{equation}
    \frac{\Lambda\a(r\p/\sqrt{2})}{\Lambda(r\p)}
    =
    \frac{2n\b (d_0\sqrt{2}/2)^3}
      {n\s (d_0/2)^3 + n\b (d_0\sqrt{2}/2)^3}
    =
    \frac{2\cdot2^{3/2}(n\b/n\s)}{1 + 2^{3/2}
    (n\b/n\s)}.
    \label{eq:correction_diluted_rp}
\end{equation}

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/occ_rate_vs_radius_model_1_brokenx.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{
        Apparent occurrence rate $\Lambda\a$, and occurrence rate $\Lambda$, 
        for a model with one type of star, one
        type of planet, and twin binaries (same as
        Figure~\ref{fig:model_1_volumes}).  We assume a twin binary
        fraction of $0.05$, and that single, primary, and secondary
        stars host planets at equal rates.  If the true planet radius
        is $r\p$, all planets detected in binaries will have apparent
        radii $r\a = r\p/\sqrt{2}$; Equation~\ref{eq:ratios} gives the
        normalizations.
    }
    \label{fig:occ_rate_model_1}
\end{figure}

We can now assess the severity of these errors, given the
binary-to-single ratio $n\b/n\s$. For stars with masses from 0.7 to
1.3~$M_\odot$, \citet{raghavan_survey_2010} found the multiplicity
fraction~--~the fraction of systems in a volume-limited sample that
are multiple~--~to be 0.44.  Assuming all multiple systems are
binaries, this means
\begin{equation}
    \frac{n\b}{n\s + n\b} \approx 0.44.
\end{equation}
Of course not all of these binaries are ``twin'' binaries as we have
assumed in our simple calculation. Perhaps only a tenth of the
binaries have pairs of stars close enough in their basic properties to
produce errors as significant as those we have been considering. In
detail, this fraction depends on the binary mass ratio distribution
and stellar mass-luminosity relation (see
Section~\ref{sec:general_formula}).  Taking a leading-order estimate
of $n\b/(n\s+n\b)\approx0.05$, we find
\begin{equation}
    \frac{\Lambda\a(r\p)}{\Lambda(r\p)} = 0.87,~~
    \frac{\Lambda\a(r\p/\sqrt{2})}{\Lambda(r\p)} = 0.26.
    \label{eq:ratios}
\end{equation}
All together, the various effects produce biases of order 10\% in the
occurrence rates.  We will see that this level of error is
characteristic of many of our more complicated models as well.

>>>>

Since the effects of binarity are most pronounced when the two stellar
components are similar, we begin by considering a universe in which
all stars are identical, with mass $M$, radius $R$, and luminosity
$L$.  Single stars are uniformly distributed in space with a number
density of $n\s$ stars per cubic parsec, and binaries are uniformly
distributed with number density $n\b$.  In this scenario, stars are
never misclassified because the combined light of a binary has the
same color and spectrum as a single star.  We further assume that all
planets have the same radius, $r$, and occur around single stars and
members of binaries at the same rate, $\Lambda(r)$.

Our naive observers conduct a transit survey. To calculate the
occurrence rate of planets with radius $r$, they count the number of
detections of signals with amplitude $(r/R)^2$.  Then they identify
all the sources that appear to have been searchable for a signal of
amplitude $(r/R)^2$.  Here and throughout the rest of this paper, we
assume that the limiting source of noise is the photon-counting noise
from the source, i.e., $\sigma \propto 1/\sqrt{F}$, where
where $F$ is the total flux of the source.
Thus the observers determine the minimum $F_0$ for which detection would have been
possible, and count the number of sources with $F_{\rm tot}>F_0$. This will
include all the single stars out to a maximum distance
\begin{equation}
  d_0 = \sqrt{\frac{L}{4\pi F_0}}.
  \label{eq:dmax}
\end{equation}
Since binaries are twice as luminous, the condition $F>F_0$ will
include binaries out to the larger distance of $d_0\sqrt{2}$.
None of the stars in binaries will
appear to have a planet of radius $r$, because of the dilution of
the transit signal.  Thus, the {\it apparent} occurrence rate $\Lambda\a$ of
planets of radius $r$ is
\begin{equation}
    \Lambda\a(r) = 
        \frac{\Lambda(r) \, n\s d_0^3}
        {n\s d_0^3 + n\b (d_0\sqrt{2})^3}.
    \label{eq:rate_at_rp}
\end{equation}
This apparent rate is larger than the true rate by a factor
\begin{equation}
    \frac{\Lambda\a(r)}{\Lambda(r)} = 
        \frac{1}{1 + 2^{3/2}(n\b/n\s)}.
    \label{eq:correction_rp}
\end{equation}

The observers will also detect some transiting planets around stars
with binary companions.  The amplitude of these signals is
\begin{equation}
  \frac{L(r/R)^2}{L + L} = \frac{1}{2}\left(\frac{r}{R}\right)^2 =
  \left(\frac{r/\sqrt{2}}{R}\right)^2,
    \label{eq:delta_obs_general} 
\end{equation}
leading the astronomers to believe they have discovered a population
of planets with radius $r/\sqrt{2}$.  To calculate the corresponding
occurrence rate, they count the stars for which this type of signal
would have been detectable.  The limiting flux for detection in this
case is $4F_0$, because the signal amplitude hase been reduced by a
factor of two and the noise level must also be reduced by a factor of
two.  The condition $F>4F_0$ is met for single stars within a distance
$d_0/2$, and binaries within a distance $d_0\sqrt{2}/2$.  Therefore,
the observers will calculate the occurrence rate of this new type of
planet to be
\begin{equation}
    \Lambda\a\left( \frac{r}{\sqrt{2}} \right) = 
        \frac{2\Lambda(r) \, n\b (d_0\sqrt{2}/2)^3}
        {n\s (d_0/2)^3 + n\b (d_0\sqrt{2}/2)^3}
    =
    \frac{2\Lambda(r)\cdot2^{3/2}(n\b/n\s)}{1 + 2^{3/2} (n\b/n\s)}.
    \label{eq:correction_diluted_rp}
\end{equation}
Figure~\ref{fig:model_1_volumes} illustrates the
volumes enclosing the apparently searchable single and binary stars.

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{figures/visualize_volumes.pdf}
    \end{center}
    \caption{ {\it Top.}---Volumes within which sources appear to be
      searchable for planets of radius $r$.  Single stars are
      searchable out to a distance $d_0$, at which point they become
      too faint to allow the detection of transits.  Binaries
      are brighter and therefore appear to be searchable out to a
      larger distance.  {\it Bottom.}---Volumes within which within
      which stars appear to be searchable for planets of radius
      $r/\sqrt{2}$.  }
    \label{fig:model_1_volumes}
\end{figure}

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/occ_rate_vs_radius_model_1_brokenx.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{ Apparent occurrence rate $\Lambda\a$, and true
    occurrence rate $\Lambda$ in a universe with only one type of
    star, one type of planet (with radius $R_\oplus$), and a number
    density of binary stars 5\% that of single stars.  The occurrence
    rate of planets with radius $R_\oplus$ is underestimated, and the
    occurrence rate of planets with radius
    $R_\oplus/\sqrt{2}$ is overestimated.}
    \label{fig:occ_rate_model_1}
\end{figure}

We can now assess the severity of the errors, for a given value of the
binary-to-single ratio $n\b/n\s$. For stars with masses from 0.7 to
1.3~$M_\odot$, \citet{raghavan_survey_2010} found the multiplicity
fraction~--~the fraction of systems in a volume-limited sample that
are multiple~--~to be 0.44.  Assuming all multiple systems are
binaries, this gives a binary fraction
\begin{equation}
  \frac{n\b}{n\s + n\b} \approx 0.44,
\end{equation}
which implies $n\b/n\s \approx 0.79$.  Of course not all of
these binaries are ``twin'' binaries as we have assumed in our simple
calculation.  Later, in Section~\ref{sec:general_formula}, we will
allow for a continuum of properties for the secondary stars. For now,
we might guess that only a tenth of the binaries have pairs of stars
close enough in brightness to produce errors as significant as those
we have been considering.  Adopting the value $n\b/n\s\approx0.05$, we
find
\begin{equation}
    \frac{\Lambda\a(r\p)}{\Lambda(r\p)} = 0.88,~~
    \frac{\Lambda\a(r\p/\sqrt{2})}{\Lambda(r\p)} = 0.25.
    \label{eq:ratios}
\end{equation}
All together, the various effects produce biases of order 10\% in the
occurrence rates.  We will see that this level of error is
characteristic of many of our more complicated models as well.
